---
title: "Project Proposal"
author: "The Repos- Jeffrey Bohrer, Alexandra Green, Anna Zhang, Kevin Lee"
format: pdf
execute: 
  warning: false
  message: false
editor: visual
---

```{r}
#| label: load-pkg-data
library(tidyverse)
library(tidymodels)
library(foreign)
# add other packages as needed

# add code to load data
```

## Introduction

...

## Data description

```{r}
autism_ds <- read.arff("Autism-Adult-Data.arff")
autism_ds <- autism_ds %>%
  rename(small_sounds = A1_Score, 
         difficult_to_understand_char  = A2_Score, 
         ease_to_read_between_lines = A3_Score ,
         focus_on_whole_picture = A4_Score,
         i_can_tell_if_someone_bored = A5_Score,
         i_can_multi_task = A6_Score,
         i_can_tell_feelings_from_faces = A7_Score,
         i_can_go_back_to_work_when_interrupted = A8_Score,
         i_like_to_collect_info_on_categories = A9_Score,
         i_find_it_hard_to_figure_out_others_intentions = A10_Score)
##Source:
## Based on AQ-10 Test:https://embrace-autism.com/aq-10/
## Score of 6 or higher is indicative of autistic traits
##UCI Machine Learning Repository. Autism Screening Adult Data Set. University of California, Irvine, 2018, https://archive.ics.uci.edu/dataset/426/autism+screening+adult.
```

...

## Exploratory data analysis

...

## Analysis approach

The clear response variable of interest is “Class/ASD,” given that it is the categorical measure of whether a test-taker possesses enough traits to be deemed in the spectrum. Considering the binary nature of the variable—taking values “Yes” or “No”—it is plausible to postulate that a logistic regression model is an optimal model for its prediction.

In terms of relevant predictor variables, the most interesting ones to add to such a model would be individual AQ-10 question responses (binary, categorical)—given that these are the main questions used to provide a final diagnosis. Additionally, considering the interaction terms between such variables along with sex, age, and ethnic background would be interesting due to the prevalence of mentions in the literature that men tend to be diagnosed more often than women (Brickhill et al., 2023), older folk tend to be underdiagnosed and under-researched (Mason et al., 2022), and that white patients are more likely to be diagnosed.

As a means of assessing such an ambitious research question, multiple models will be fitted until arriving the most parsimonious one and assessing whether they have a statistically significant effect on the final diagnosis as provided by the app. 


## Data dictionary

The data dictionary can be found [here](https://archive.ics.uci.edu/dataset/426/autism+screening+adult) 

## References
Brickhill, Rae et al. “Autism, thy name is man: Exploring implicit and explicit gender bias in autism perceptions.” PloS one vol. 18,8 e0284013. 23 Aug. 2023, doi:10.1371/journal.pone.0284013

Mason, David et al. “Older Age Autism Research: A Rapidly Growing Field, but Still a Long Way to Go.” Autism in adulthood : challenges and management vol. 4,2 (2022): 164-172. doi:10.1089/aut.2021.0041

Aylward, Brandon S et al. “Racial, Ethnic, and Sociodemographic Disparities in Diagnosis of Children with Autism Spectrum Disorder.” Journal of developmental and behavioral pediatrics : JDBP vol. 42,8 (2021): 682-689. doi:10.1097/DBP.0000000000000996

Thabtah, Fadi, et al. “A New Computational Intelligence Approach to Detect Autistic Features for Autism Screening.” Applied Soft Computing, vol. 83, 2019, 105748. ScienceDirect, https://doi.org/10.1016/j.asoc.2018.06.011.



